Proof-Reconstruction in Type Theory for Propositional Logic

### 2017-10-22 paper/sections/proof-reconstruction/resolve.tex
[ paper ] working on resolve.

Logic equivalence between propositions is a major issue to justify
prover reasoning steps. Since we left out semantics to treat only the
syntax aspects of the propositions, our approach shows equivalence
by converting propositions to their conjunctive normal forms,
and reordering those and the inner disjunctions to match them.
In this section, we provide some theorems to perform such reordering tasks,
the goals are the following.

\begin{itemize}
  \item Reorder a disjunction $φ = φ₁ ∨ φ₂ ∨ \cdots ∨ φₙ$ to match with
$ψ = ψ₁ ∨ ψ₂ ∨ \cdots ∨ ψₙ$. For all $i$, there exists at least one $j$ such
that $φᵢ ≡ ψⱼ$.
  \item Reorder a CNF formula, $φ = φ₁ ∧ φ₂ ∧ \cdots$, to match with another
CNF formula $ψ = ψ₁ ∧ ψ₂ ∧ \cdots$. For all $i$, there exists at least one $j$
such that $φᵢ ≡ ψⱼ$.
\end{itemize}

First, we define the $\rm{rdisj}$ function to convert
a disjunction in its right-associative form.
The $\rm{rconj}$ function that converts conjunction to its right-associative form
is not presented here, but is defined in a similar way (see its definition
in~\cite{AgdaMetis}).

\begin{definition}[rdisj]
\label{def:rdisj}
  \begin{align*}
    \begin{split}
    \label{eq:rdisj-def}
      &\rm{rdisj}(φ) : \Prop \to \Prop\\
      &\rm{rdisj}(φ) =
      \begin{cases}
      \rm{rdisj}(φ₁ ∨ (φ₂ ∨ φ₃)),  & \text{ if } φ ≡ (φ₁ ∨ φ₂) ∨ φ₃\\
      φ₁ ∨ \rm{rdisj}(φ₂),         & \text{ if } φ ≡ φ₁ ∨ φ₂\\
      φ,                      & \text{ otherwise.}
      \end{cases}
    \end{split}
  \end{align*}
\end{definition}

\begin{mainlemma}[lem-rdisj]
\label{lem:lem-rdisj}
  If $Γ ⊢ φ$ then $Γ ⊢ \rm{rdisj}(φ)$.
\end{mainlemma}

\begin{definition}[build$_{\boldsymbol{∨}}$]
\begin{equation*}
  \begin{split}
  \label{eq:build-or}
  &\rm{build}_{∨} : \Prop → \Prop → \Prop\\
  &\rm{build}_{∨}(φ, ψ) =%\\
  % &\hspace{3mm}
  \begin{cases}
  ψ, &\text{ if } φ ≡ ψ\\
  ψ, &\text{ if } ψ ≡ ψ₁ ∨ ψ₂\text{, } \rm{build}_{∨}(φ, ψ_{i}) ≡ ψ_{i}\text{ for some }i=1, 2\\
  φ, &\text{ otherwise.}
  \end{cases}
  \end{split}
\end{equation*}
\end{definition}

@@

\begin{mainlemma}[lem-build$_{\boldsymbol{∨}}$]
\label{lem:lem-build-or}
If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \rm{build}_{∨}(φ)$.
\end{mainlemma}

From now, we assume all propositions to be right-associative unless
otherwise stated.
The $\ffactor$ function in Eq.~(\ref{eq:factor-definition}) intends to
simplify a special case of disjunction, the repeated disjuncts
(e.g., $\ffactor(φ ∨ φ) \mapsto\ φ$).
Notice that other cases like $φ ∨ (ψ ∨ φ)$ do no reduce to $(ψ ∨ φ)$.

\begin{definition}[factor]
\begin{align}
\begin{split}
  \label{eq:factor-definition}
  &\ffactor : \Prop → \Prop\\
  &\ffactor(φ) =
  \begin{cases}
    φ₁,  &\text{ if }φ ≡ φ₁ ∨ φ₂ \text{, } φ₁ ≡ \ffactor(φ₂)\\
    φ,   &\text{ otherwise.}
  \end{cases}
\end{split}
\end{align}
\end{definition}

@@

To construct a disjunction $ψ$ from a disjunction $φ$, we have used ideas
from the description in \citeauthor{bohme2010}~\cite{bohme2010} to prove
equality between nested disjunctions.
The result is the $\rm{sbuild}_{∨}$ function
that uses every disjunct from the source formula, $φ$, to build up the
target disjunction $ψ$.

\begin{definition}[sbuild$_{\boldsymbol{∨}}$]
\begin{align*}
  \begin{split}
  \label{eq:strong-build-or-definition}
    &\rm{sbuild}_{∨} : \Prop → \Prop → \Prop\\
    &\rm{sbuild}_{∨}(φ, ψ) =%\\
    % &\hspace{3mm}
    \begin{cases}
      \ffactor(\rm{build}_{∨}(φ₁, ψ) ∨ \rm{build}_{∨}(φ₂,ψ)),
      &\text{ if }φ≡ φ₁ ∨ φ₂\\
      \rm{build}_{∨}(φ, ψ),  & \text{ otherwise.}
    \end{cases}
  \end{split}
\end{align*}
\end{definition}

\begin{mainlemma}[lem-sbuild$_{\boldsymbol{∨}}$]
\label{lem:lem-sbuild-or}
If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \rm{sbuild}_{∨}(φ, ψ)$.
\end{mainlemma}

@@

\begin{definition}[reorder$_{\boldsymbol{∨}}$]
\begin{align*}
    \label{eq:reorder-or}
    \begin{split}
    &\freorder_{∨} : \Prop \to \Prop \to \Prop\\
    &\freorder_{∨} (φ, ψ) = \rm{sbuild}_{∨}(rdisj(φ), ψ).
    \end{split}
  \end{align*}
\end{definition}

\begin{mainth}[thm-reorder$_{\boldsymbol{∨}}$]
  \label{thm-reorder-or}
  If $Γ ⊢ φ$ and $ψ: \Prop$ then $Γ ⊢ \freorder_{∨}(φ, ψ)$.
\end{mainth}

\begin{proof}
Use Lemma~\ref{lem:lem-rdisj} and Lemma~\ref{lem:lem-sbuild-or}.
\end{proof}

@@

\begin{mainth}[thm-reorder$_{\boldsymbol{∧∨}}$]
  \label{thm:thm-reorder-and-or}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \freorder_{∧∨}(φ, ψ)$.
\end{mainth}

@@

\begin{mainth}[thm-resolve]
  \label{thm:thm-resolve}
  If $Γ ⊢ φ₁$, $Γ ⊢ φ₂$, a literal $l$, and $ψ : \Prop$ then
  $Γ ⊢ \fresolve(φ₁, φ₂, l, ψ)$.
\end{mainth}

@@

\begin{figure}
\label{fig:resolve-example}
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ ¬ r ∨ p ∨ q$}
  \AxiomC{$Γ ⊢ p ∨ q ∨ r $}
  \RightLabel{thm-resolve}
  \BinaryInfC{$Γ ⊢ \fresolve(¬ r ∨ p ∨ q, p ∨ q ∨ r, r, p ∨ q)$}
  \UnaryInfC{$Γ ⊢ p ∨ q$}
  \end{bprooftree}
\end{equation*}
\caption{Example of a resolution in \Metis using
Thm.~\ref{thm:thm-resolve}.}
%Application of Thm.~\ref{thm:thm-resolve} for the \TSTP excerpt Fig.~\ref{fig:resolve-tstp-example}.}
\end{figure}

>>>>

Logic equivalence between propositions is a major issue to justify
prover reasoning steps. Since we left out semantics to treat only the
syntax aspects of the propositions, our approach shows equivalence
by converting propositions to their conjunctive normal forms,
and reordering those and the inner disjunctions to match them.
Below, we provide some lemmas to perform such reordering tasks,
we omit the proofs but they can found in~\cite{AgdaMetis}.

% \begin{itemize}
%   \item Reorder a disjunction $φ = φ₁ ∨ φ₂ ∨ \cdots ∨ φₙ$ to match with
% $ψ = ψ₁ ∨ ψ₂ ∨ \cdots ∨ ψₙ$. For all $i$, there exists at least one $j$ such
% that $φᵢ ≡ ψⱼ$.
%   \item Reorder a CNF formula, $φ = φ₁ ∧ φ₂ ∧ \cdots$, to match with another
% CNF formula $ψ = ψ₁ ∧ ψ₂ ∧ \cdots$. For all $i$, there exists at least one $j$
% such that $φᵢ ≡ ψⱼ$.
% \end{itemize}

First, we define \rm{rassoc}$_{⬓}$ function in Eq.~(\ref{def:right-square})
to convert a disjunctions or a conjunction into its right-associative form.
We use $\rm{rassoc}_{∧}$ to convert conjunctions and $\rm{rassoc}_{∨}$ for
disjunctions.

\begin{equation}
\label{def:right-square}
  \begin{aligned}
    &\hspace{.495mm}\rm{rassoc}_{⬓} : \Prop \to \Prop\\
    &\begin{array}{lll}
    \rm{rassoc}_{⬓} &((φ₁ ⬓ φ₂) ⬓ φ₃) &= \rm{rassoc}_{⬓}~(φ₁~⬓~(φ₂ ⬓ φ₃))\\
    \rm{rassoc}_{⬓} &(φ₁ ⬓ φ₂)        &= φ₁~∨~\rm{rassoc}_{⬓}~φ₂\\
    \rm{rassoc}_{⬓} &φ                &= φ
    \end{array}
  \end{aligned}
\end{equation}

\begin{example}\hspace{5mm}\\
\begin{enumerate}
\item $\rm{rassoc}_{∨}~(((p ∨ ((q ∨ r) ∨ s)) ∨ t) = p ∨ (q ∨ (r ∨ (s ∨ t)))$.
\item $\rm{rassoc}_{∧}~((((p ∧ q) ∧ r) ∧ s) ∧ t) = p ∧ (q ∧ (r ∧ (s ∧ t)))$.
\end{enumerate}
\end{example}

\begin{remark}
In \TPTP syntax, the formulas are in left-associative form by default.
Despite of that convention, \Metis understands a formula to be in right-associative form by default. As we mentioned above, this is a matter to take into account for
the proof-reconstruction. We opted to work as far as possible with right-associative formulas in~\cite{AgdaMetis}.
\end{remark}

\begin{mainlemma}
\label{lem:lem-rassoc}
  If $Γ ⊢ φ$ then $Γ ⊢ \rm{rassoc}_{⬓}~φ$.
\end{mainlemma}

\begin{equation*}
  \begin{split}
  \label{eq:build-or}
  &\rm{build}_{∨} : \Prop → \Prop → \Prop\\
  &\rm{build}_{∨}(φ, ψ) =%\\
  % &\hspace{3mm}
  \begin{cases}
  ψ, &\text{ if } φ ≡ ψ\\
  ψ, &\text{ if } ψ ≡ ψ₁ ∨ ψ₂\text{, } \rm{build}_{∨}(φ, ψ_{i}) ≡ ψ_{i}\text{ for some }i=1, 2\\
  φ, &\text{ otherwise.}
  \end{cases}
  \end{split}
\end{equation*}

@@

\begin{mainlemma}[lem-build$_{\boldsymbol{∨}}$]
\label{lem:lem-build-or}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \rm{build}_{∨}~φ~ψ$.
\end{mainlemma}

From now, we assume all propositions to be right-associative unless
otherwise stated.
The $\ffactor$ function in Eq.~(\ref{eq:factor-definition})
simplifies a special case of disjunction, the repeated disjuncts
(e.g., $\ffactor(φ ∨ φ) = φ$).
Notice that other cases like $φ ∨ (ψ ∨ φ)$ do no reduce to $ψ ∨ φ$.

\begin{align}
\begin{split}
  \label{eq:factor-definition}
  &\ffactor : \Prop → \Prop\\
  &\ffactor(φ) =
  \begin{cases}
    φ₁,  &\text{ if }φ ≡ φ₁ ∨ φ₂ \text{ and } φ₁ ≡ \ffactor(φ₂)\\
    φ,   &\text{ otherwise.}
  \end{cases}
\end{split}
\end{align}

@@

To construct a disjunction $ψ$ from a disjunction $φ$, we have used ideas
from the description in \citeauthor{bohme2010}~\cite{bohme2010} to prove
equality between nested disjunctions.
We define the $\rm{sbuild}_{∨}$ function in
Eq.~(\ref{eq:strong-build-or-definition}).
that uses every disjunct from the source formula, $φ$, to build up the
target disjunction $ψ$.

\begin{align}
  \begin{split}
  \label{eq:strong-build-or-definition}
    &\rm{sbuild}_{∨} : \Prop → \Prop → \Prop\\
    &\rm{sbuild}_{∨}(φ, ψ) =%\\
    % &\hspace{3mm}
    \begin{cases}
      \ffactor(\rm{build}_{∨}(φ₁, ψ) ∨ \rm{build}_{∨}(φ₂,ψ)),
      &\text{ if }φ≡ φ₁ ∨ φ₂\\
      \rm{build}_{∨}(φ, ψ),  & \text{ otherwise.}
    \end{cases}
  \end{split}
\end{align}

\begin{mainlemma}
\label{lem:lem-sbuild-or}
If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \rm{sbuild}_{∨}(φ, ψ)$.
\end{mainlemma}

@@

\begin{mainth}[thm-reorder$_{\boldsymbol{∨}}$]
  \label{thm-reorder-or}
  If $Γ ⊢ φ$ and $ψ: \Prop$ then $Γ ⊢ \freorder_{∨}~φ~ψ~$, where,
  \begin{align*}
    \label{eq:reorder-or}
    \begin{split}
    &\freorder_{∨} : \Prop \to \Prop \to \Prop\\
    &\freorder_{∨}~φ~ψ~= \rm{sbuild}_{∨}~(rassoc_{∨}~φ)~ψ.
    \end{split}
  \end{align*}
\end{mainth}

\begin{proof}
Use Lemma~\ref{lem:lem-rassoc} and Lemma~\ref{lem:lem-sbuild-or}.
\end{proof}

@@

\begin{mainth}[thm-reorder$_{\boldsymbol{∧∨}}$]
  \label{thm:reorder-and-or}
  If $Γ ⊢ φ$ and $ψ : \Prop$ then $Γ ⊢ \freorder_{∧∨}(φ, ψ)$.
\end{mainth}

@@

\begin{mainth}[thm-resolve]
  \label{thm:resolve}
  If $Γ ⊢ φ₁$, $Γ ⊢ φ₂$, a literal $l$, and $ψ : \Prop$ then
  $Γ ⊢ \fresolve(φ₁, φ₂, l, ψ)$.
\end{mainth}

@@

\begin{figure}
\label{fig:resolve-example}
\begin{equation*}
  \begin{bprooftree}
  \AxiomC{$Γ ⊢ ¬ r ∨ p ∨ q$}
  \AxiomC{$Γ ⊢ p ∨ q ∨ r $}
  \RightLabel{thm-resolve}
  \BinaryInfC{$Γ ⊢ \fresolve(¬ r ∨ p ∨ q, p ∨ q ∨ r, r, p ∨ q)$}
  \UnaryInfC{$Γ ⊢ p ∨ q$}
  \end{bprooftree}
\end{equation*}
\caption{Example of a resolution in \Metis using
Thm.~\ref{thm:resolve}.}
%Application of Thm.~\ref{thm:resolve} for the \TSTP excerpt Fig.~\ref{fig:resolve-tstp-example}.}
\end{figure}
